Malacology: A Programmable Storage System

### 2017-02-22 paper/paper.tex
paper: version 1 - adds future work and clarifies metadata - brings in more citations and all our projects - still under the page limit (we get 14 + refs)
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Uncomment for diff
%\includecomment{notes}
%\newcommand{\newcomment}[1]{{\textcolor{red}{\textbf{#1}}}}
%\newcommand{\oldcomment}[1]{{\textcolor{blue}{\textbf{\sout{#1}}}}}
%\newcommand{\addressesissue}[1]{{\textcolor{red}{{\it (Addresses issue(s): {#1})}}}}

% Uncomment for camera-ready
\excludecomment{notes}
\newcommand{\newcomment}[1]{{\textcolor{black}{#1}}}
\newcommand{\oldcomment}[1]{}
\newcommand{\addressesissue}[1]{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract} Storage systems are caught between the need to evolve data
processing applications efficiently and quickly, and the increasing velocity
with which storage device technology evolves. This puts tremendous pressure on
storage systems to support rapid change both in terms of their interfaces and
their performance. But adapting storage systems can be difficult because
unprincipled changes might jeopardize years of code-hardening and performance
optimization efforts that were necessary for users to entrust their data to the
storage system. We introduce the concept of a programmable storage system, a
system that exposes internal services and abstractions with the goal of
re-using and building new services on top of the storage stack. We present
Malacology, a prototype implementation of a programmable storage system to
explore how existing abstractions of common storage system services can be
leveraged to adapt to the needs of new data processing systems and the
increasing variety of storage devices. We show how this approach allows a large
degree of flexibility for storage systems to evolve without sacrificing the
robustness of their code-hardened subsystems. We illustrate the advantages and
challenges of programmability by composing existing primitives into two new
higher-level services: a metadata load balancer and a high-performance
distributed shared-log. The evaluation demonstrates that our services inherit
desirable qualities of the back-end storage system, including the ability to
balance load, efficiently propagate cluster metadata, recover from failure, and
to navigate trade-offs between latency and throughput using leases.
\end{abstract}

@@

{\it \textbf{Contribution 1:}} We define a programmable storage system to be a
storage system that facilitates the re-use and extension of existing storage
abstractions provided by the underlying software stack, to enable the creation
of new services via composition.  A programmable storage system can be realized
by exposing existing functionality (such as metadata services and
synchronization and monitoring capabilities) as interfaces that can be ``glued
together'' in a variety of ways using a high-level language. Programmable
storage differs from \emph{active storage}~\cite{riedel:vldb98}---the injection
and execution of code within a storage system or storage device---in that the
former is applicable to any component of the storage system, while the latter
focuses at the data access level. Given this contrast, we can say that active
storage is an example of how one internal component (the storage layer) is
exposed in a programmable storage system.

To illustrate the benefits and challenges of this approach we have designed and
evaluated Malacology, a programmable storage system that facilitates the
construction of new services by re-purposing existing subsystem abstractions of
the storage stack.  We build Malacology in Ceph, a popular open source software
storage stack.  We choose Ceph to demonstrate the concept of programmable
storage because it offers a broad spectrum of existing services, including
distributed locking and caching services provided by metadata servers,
durability and object interfaces provided by the backend object store, and
propagation of consistent cluster state provided by the monitoring service (see
Figure~\ref{fig:overview}).  Malacology is expressive enough to provide the
functionality necessary for implementing new services.

@@

% straw man example
Keeping track of state in a distributed system is an essential part of any
successful service and a necessary component in order to diagnose and detect
failures, when they occur. This is further complicated by variable propagation
delays and heterogeneous hardware in dynamic environments.

\newcomment{\\ \noindent\it{\textbf{Existing Ceph Implementation: }}}
In the case of Ceph, a consistent view of cluster state among server daemons
and clients is critical to provide strong consistency guarantees to clients.
Ceph maintains cluster state information in per-subsystem ({\it e.g.} the
metadata server) data structures called ``maps'' that record membership and
status information.  A Paxos~\cite{lamport_parttime_1998} monitoring service is
responsible for integrating state changes into cluster maps, responding to
requests from out-of-date clients and synchronizing members of the cluster
whenever there is a change in a map so that they all observe the same system
state. As a fundamental building block of many system designs, consensus
abstractions such as Paxos are a common technique for maintaining consistent
data versions, and are a useful system to expose.

The default behavior of the monitor can be seen as a Paxos-based notification
system, similar to the one introduced in~\cite{burrows_chubby_2006}, allowing
clients to identify when new values (termed epochs in Ceph) are associated to
given maps.  Since Ceph does not expose this service directly, as part of our Malacology implementation, we expose a key-value
service designed for managing configuration metadata that is built on top of
the consensus engine. Since the monitor is intended to be out of the
high-performance I/O path, a general guideline is to make use of this
functionality infrequently and to assign small values to maps.\\

\noindent{\it{\textbf{Malacology: }}} Malacology exposes a strongly-consistent view of
time-varying service metadata as a service rather than a hidden internal
component. Malacology provides a generic API for adding arbitrary values to
existing subsystem cluster maps. As a consequence of this, applications can
define simple but useful service-specific logic to the strongly-consistent
interface, such as authorization control (just specific clients can write new
values) or triggering actions based on specific values (e.g. sanitize values).
The higher level services we implement in \S\ref{sec:services} make use of this
functionality to register, version and propagate dynamic code (Lua scripts) for
new object interfaces defined in storage daemons (\S\ref{object-data-interface}) and
policies in the load balancer \S\ref{malacology:mds}.  Using this service
guarantees that interface definitions are not only made durable, but are
transparently and consistently propagated throughout the cluster so that
clients are properly synchronized with the latest interfaces.\\

@@

\begin{figure}[tbp]
\centering
\includegraphics{figures/implementation.png}
\caption{Malacology allows users to dynamically define object/metadata
interfaces by extending the object storage daemon (OSD) and metadata server
(MDS) subsystems with an embedded Lua VM.  It uses the service metadata
interface (M) to propagate interfaces/versions across the cluster
\label{fig:implementation}}
\end{figure}

@@

\oldcomment{\it{\textbf{Malacology:}} Interfaces are added in strategic spots in the
metadata service for guarding shared resources, defining new file types,
and load balancing.}

@@

In general, file system metadata servers manages client sessions, allowing
clients to obtain locks (e.g. file byte ranges), and capabilities (e.g. to
cache file data). Clients and metadata servers use a cooperative protocol in
which clients voluntarily release resources back to the metadata service in
order to implement sharing policies.

\newcomment{\\ \noindent\it{\textbf{Existing Ceph Implementation: }}}
In Ceph, the locking service implements a capability-based system that
expresses what data and metadata clients are allowed to access as well as what
state they may cache and modify locally.  While designed for the file
abstraction, indexing, locking, and caching are all common services that are
useful to a broad spectrum of applications.  Distributed applications that
share centralized resources (e.g. a database or directory) face similar
challenges which are often solved using application-specific sharding.

@@

Applications that manage large amounts of metadata (e.g. users or database
snapshots) often require a naming service.  The metadata service manages a
POSIX file system hierarchy where files and directories are represented as
inode data structures and expose a POSIX file interface.

@@

\newcomment{Many large scale storage systems separate metadata and data IO so
that the corresponding services can scale independently. Metadata requests
transfer small amounts of data and they happen relatively frequently so many
systems employ separate metadata clusters.}

\newcomment{\\ \noindent {\it{\textbf{Existing Ceph Implementation: }}}} Ceph
also addresses the challenge of balancing metadata load with a separate
metadata cluster. This cluster uses load balancing policies to migrate
directory inodes around the cluster to alleviate overload at single
nodes~\cite{weil:sc2004-dyn-metadata}.  The policies use metrics based on
system state (e.g.  CPU and memory utilization) and statistics collected by the
cluster (e.g. the popularity of an inode). Ceph uses dynamic subtree
partitioning to move variable sized namespace subtrees. These units can be
shipped anywhere (i.e., to any metadata server of any capacity) at any time for
any reason. The original balancer was designed with hard-coded policies and
tunables.\\

\noindent {\it{\textbf{Malacology: }}}
Mantle~\cite{sevilla:sc15-mantle}, a programmable metadata load balancer, has
been implemented on Malacology so that its infrastructure can enjoy all the
properties of existing internal abstractions.  The load balancing logic is
intact but the infrastructure has been improved to safely load, manage, and
persist its load balancing policies.  Composing Mantle with the service
metadata and durability interfaces, Mantle can safely version balancer
policies, save balancer policies in the backend object store and centralize
warnings/errors. When combined with file types, load balancing policies can
express a variety of rules such as multi-tenant and multi-application
balancing.

@@

\noindent {\it{\textbf{Malacology: }}} Metadata service policies and object storage
interfaces are stored durability within RADOS and managed by storing references
with the object maps. Since the cluster already propagates a consistent view of
these data structures, we use this service to automatically install interfaces
in OSDs, and install policies within the metadata server daemon (MDS) such that clients and daemons are
synchronized on correct implementations without restarting.

@@

@@

Another view of programmable storage systems is one of tailoring systems
resources to applications~\cite{arpaci:sosp01}. Related efforts include the
Exokernel~\cite{engler:sosp95}, SPIN~\cite{bershad:sosp95} and
Vino~\cite{seltzer:osdi96} projects; the latter two addressed the ability of injecting
code into the kernel to specialize resource management. Another approach is to
pass hints between the different layers of the IO stack to bridge the semantic
gap between applications and
storage~\cite{arpaci:sosp01,sivathanu:fast03,mesnier:sosp11}).

@@

\section{Conclusion and Future Work}
\label{conclusion-and-future-work}

@@

Future work will focus on constructing more interfaces to support a wide
variety of storage system services that can be configured on-the-fly in existing
systems. This work is one point along that path to producing general-purpose
storage systems that can
target  special-purpose applications.  Ultimately we want to utilize
declarative methods for expressing new services.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Uncomment for diff 0
%\includecomment{notes}
%\newcommand{\newcomment}[1]{{\textcolor{red}{\textbf{#1}}}}
%\newcommand{\oldcomment}[1]{{\textcolor{blue}{\textbf{\sout{#1}}}}}
%\newcommand{\addressesissue}[1]{{\textcolor{red}{{\it (Addresses issue(s): {#1})}}}}

% Uncomment for diff 1
\includecomment{notes}
\newcommand{\newcommentone}[1]{{\textcolor{red}{\textbf{#1}}}}
\newcommand{\oldcommentone}[1]{{\textcolor{blue}{\textbf{\sout{#1}}}}}
\newcommand{\newcomment}[1]{{\textcolor{black}{#1}}}
\newcommand{\oldcomment}[1]{}
\newcommand{\addressesissue}[1]{{\textcolor{red}{{\it (Addresses issue(s): {#1})}}}}

% Uncomment for camera-ready
%\excludecomment{notes}
%\newcommand{\newcomment}[1]{{\textcolor{black}{#1}}}
%\newcommand{\oldcomment}[1]{}
%\newcommand{\newcommentone}[1]{{\textcolor{black}{#1}}}
%\newcommand{\oldcommentone}[1]{}
%\newcommand{\addressesissue}[1]{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract} Storage systems are caught between the need to evolve data
processing applications efficiently and quickly, and the increasing velocity
with which storage device technology evolves. This puts tremendous pressure on
storage systems to support rapid change both in terms of their interfaces and
their performance. But adapting storage systems can be difficult because
unprincipled changes might jeopardize years of code-hardening and performance
optimization efforts that were necessary for users to entrust their data to the
storage system. We introduce the concept of a programmable storage system, a
system that exposes internal services and abstractions with the goal of
re-using and building new services on top of the storage stack. We present
Malacology, a prototype implementation of a programmable storage system to
explore how existing abstractions of common storage system services can be
leveraged to adapt to the needs of new data processing systems and the
increasing variety of storage devices. We show how this approach allows a large
degree of flexibility for storage systems to evolve without sacrificing the
robustness of their code-hardened subsystems. We illustrate the advantages and
challenges of programmability by composing existing primitives into two new
higher-level services: a \newcommentone{file system} metadata load balancer and a high-performance
distributed shared-log. The evaluation demonstrates that our services inherit
desirable qualities of the back-end storage system, including the ability to
balance load, efficiently propagate cluster metadata, recover from failure, and
to navigate trade-offs between latency and throughput using leases.
\addressesissue{4}
\end{abstract}

@@

{\it \textbf{Contribution 1:}} We define a programmable storage system to be a
storage system that facilitates the re-use and extension of existing storage
abstractions provided by the underlying software stack, to enable the creation
of new services via composition.  A programmable storage system can be realized
by exposing existing functionality (such as \newcommentone{file system and cluster} metadata services and
synchronization and monitoring capabilities) as interfaces that can be ``glued
together'' in a variety of ways using a high-level language. Programmable
storage differs from \emph{active storage}~\cite{riedel:vldb98}---the injection
and execution of code within a storage system or storage device---in that the
former is applicable to any component of the storage system, while the latter
focuses at the data access level. Given this contrast, we can say that active
storage is an example of how one internal component (the storage layer) is
exposed in a programmable storage system.
\addressesissue{4}

To illustrate the benefits and challenges of this approach we have designed and
evaluated Malacology, a programmable storage system that facilitates the
construction of new services by re-purposing existing subsystem abstractions of
the storage stack.  We build Malacology in Ceph, a popular open source software
storage stack.  We choose Ceph to demonstrate the concept of programmable
storage because it offers a broad spectrum of existing services, including
distributed locking and caching services provided by \newcommentone{file system} metadata servers,
durability and object interfaces provided by the backend object store, and
propagation of consistent cluster state provided by the monitoring service (see
Figure~\ref{fig:overview}).  Malacology is expressive enough to provide the
functionality necessary for implementing new services.
\addressesissue{4}

@@

% straw man example
Keeping track of state in a distributed system is an essential part of any
successful service and a necessary component in order to diagnose and detect
failures, when they occur. This is further complicated by variable propagation
delays and heterogeneous hardware in dynamic environments.
\newcommentone{Service metadata is information about the daemons in the system
and includes membership details, hardware layout ({\it e.g.}, racks, power
supplies, etc.), data layout, and daemon state/configuration. It differs from
traditional file system metadata which is information about files. For the rest
of the paper when we refer to metadata server or service, we mean a daemon that manages file
system metadata.}
\addressesissue{4}

\newcomment{\\ \noindent\it{\textbf{Existing Ceph Implementation: }}} In the
case of Ceph, a consistent view of cluster state among server daemons and
clients is critical to provide strong consistency guarantees to clients.  Ceph
maintains cluster state information in per-subsystem  data structures called
``maps'' that record membership and status information.  A
Paxos~\cite{lamport_parttime_1998} monitoring service is responsible for
integrating state changes into cluster maps, responding to requests from
out-of-date clients and synchronizing members of the cluster whenever there is
a change in a map so that they all observe the same system state. As a
fundamental building block of many system designs, consensus abstractions such
as Paxos are a common technique for maintaining consistent data versions, and
are a useful system to expose.

The default behavior of the monitor can be seen as a Paxos-based notification
system, similar to the one introduced in~\cite{burrows_chubby_2006}, allowing
clients to identify when new values (termed epochs in Ceph) are associated to
given maps.  Since Ceph does not expose this service directly, as part of our Malacology implementation, we expose a key-value
service designed for managing \oldcommentone{configuration} \newcommentone{service} metadata that is built on top of
the consensus engine. Since the monitor is intended to be out of the
high-performance I/O path, a general guideline is to make use of this
functionality infrequently and to assign small values to maps.\\

\noindent{\it{\textbf{Malacology: }}} Malacology exposes a strongly-consistent view of
time-varying service metadata as a service rather than a hidden internal
component. Malacology provides a generic API for adding arbitrary values to
existing subsystem cluster maps. As a consequence of this, applications can
define simple but useful service-specific logic to the strongly-consistent
interface, such as authorization control (just specific clients can write new
values) or triggering actions based on specific values (e.g. sanitize values).
The higher-level services we implement in \S\ref{sec:services} make use of this
functionality to register, version and propagate dynamic code (Lua scripts) for
new object interfaces defined in storage daemons (\S\ref{object-data-interface}) and
policies in the load balancer \S\ref{malacology:mds}.  Using this service
guarantees that interface definitions are not only made durable, but are
transparently and consistently propagated throughout the cluster so that
clients are properly synchronized with the latest interfaces.\\

@@

\begin{figure}[tbp]
\centering
\includegraphics{figures/implementation.png}
\caption{Malacology allows users to dynamically define object/\newcommentone{file system} metadata
interfaces by extending the object storage daemon (OSD) and metadata server
(MDS) subsystems with an embedded Lua VM.  It uses the service metadata
interface (M) to propagate interfaces/versions across the cluster
\label{fig:implementation}}
\end{figure}

@@

\oldcomment{\it{\textbf{Malacology:}} Interfaces are added in strategic spots in the
\newcommentone{file system} metadata servers for guarding shared resources, defining new file types,
and load balancing.}

@@

In general, file system metadata servers manages client sessions, allowing
clients to obtain locks (e.g. file byte ranges), and capabilities (e.g. to
cache file data). Clients and metadata servers use a cooperative protocol in
which clients voluntarily release resources back to the \newcommentone{file system} metadata service in
order to implement sharing policies.

\newcomment{\\ \noindent\it{\textbf{Existing Ceph Implementation: }}}
In Ceph, the locking service implements a capability-based system that
expresses what data and \newcommentone{file system} metadata clients are allowed to access as well as what
state they may cache and modify locally.  While designed for the file
abstraction, indexing, locking, and caching are all common services that are
useful to a broad spectrum of applications.  Distributed applications that
share centralized resources (e.g. a database or directory) face similar
challenges which are often solved using application-specific sharding.

@@

Applications that manage large amounts of \newcommentone{file system} metadata (e.g. users or database
snapshots) often require a naming service.  The metadata service manages a
POSIX file system hierarchy where files and directories are represented as
inode data structures and expose a POSIX file interface.

@@

\newcomment{Many large scale storage systems separate file system metadata and data IO so
that the corresponding services can scale independently. Metadata requests
transfer small amounts of data and they happen relatively frequently so many
systems employ separate file system metadata clusters.}

\newcomment{\\ \noindent {\it{\textbf{Existing Ceph Implementation: }}}} Ceph
also addresses the challenge of balancing \newcommentone{file system} metadata load with a separate
metadata cluster. This cluster uses load balancing policies to migrate
directory inodes around the cluster to alleviate overload at single
nodes~\cite{weil:sc2004-dyn-metadata}.  The policies use metrics based on
system state (e.g.  CPU and memory utilization) and statistics collected by the
cluster (e.g. the popularity of an inode). Ceph uses dynamic subtree
partitioning to move variable sized namespace subtrees. These units can be
shipped anywhere (i.e., to any metadata server of any capacity) at any time for
any reason. The original balancer was designed with hard-coded policies and
tunables.\\

\noindent {\it{\textbf{Malacology: }}}
Mantle~\cite{sevilla:sc15-mantle}, a programmable \newcomment{file system}metadata load balancer, has
been implemented on Malacology so that its infrastructure can enjoy all the
properties of existing internal abstractions.  The load balancing logic is
intact but the infrastructure has been improved to safely load, manage, and
persist its load balancing policies.  Composing Mantle with the service
metadata and durability interfaces, Mantle can safely version balancer
policies, save balancer policies in the backend object store and centralize
warnings/errors. When combined with file types, load balancing policies can
express a variety of rules such as multi-tenant and multi-application
balancing.

@@

\noindent {\it{\textbf{Malacology: }}} Metadata service policies and object storage
interfaces are stored durability within RADOS and managed by storing references
with the object maps. Since the cluster already propagates a consistent view of
these data structures, we use this service to automatically install interfaces
in OSDs, and install policies within the metadata server daemon \oldcommentone{(MDS)} such that clients and daemons are
synchronized on correct implementations without restarting.

@@

\section{Future Work}

\newcommentone{This work is one point along the path towards producing
general-purpose storage systems that can target special-purpose applications.
Continuing to expore existing storage system functionality as reusable building
blocks will enable the proliferation of storage-centric APIs, allowing
developers to custom-fit their applications to take advantage of the
capabilities of the underlying system. When adding new interfaces and composing
them together, careful consideration should be made to address security, access
control, and safety. Correctness and performance will surely be affected if
interfaces are combined poorly and these drawbacks will be quantified.}
\addressesissue{5}

\newcommentone{The next higher-level services we plan to build using the
interfaces in Table~\ref{table:examples} are: an elastic cloud database, a data
layout manager, and a data processing engine. Some cloud databases aggressively
cache data to hide the overheads and weaker consistency semantics of the
distributed file systems they sit on~\cite{pavlo:sigmod16} but using Malacology
we could build a database that works well because of the storage system not in
spite of it. We plan to use the data IO interface for push down predicates, the
file type interface for maintaining indices and stats, and the durability
interface internals for ingest and data movement. The data layout manager would
be similar to DRepl~\cite{ionkov:msst2013} and would use the file type,
durability, and data I/O interface to manage different views of data. Finally,
the data processing engine would be similar to MapReduce and would use the data
I/O, shared resource, and load balancing interface to implement the resource
manager, computation engine, and storage platform in the Hadoop ecosystem.}

\newcommentone{ Our experience with Zlog and Mantle illustrates that the labor
of wrapping existing services in interfaces is justified by the power and
flexibility they afford to programmers.  In exchange for this flexibility,
however, programmable storage systems force programmers to forfeit the
\emph{protection from change} afforded by narrow storage interfaces like the
POSIX API.  To implement applications built atop systems such as Malacology,
programmers must navigate a complex design space, simultaneously addressing
often orthogonal concerns including functional correctness, performance and
fault tolerance. Worse still, these compositions can be sensitive to changes in
the underlying environment (such as hardware upgrades, software version changes
and evolving workloads).  For example, a major version change in Ceph required
us to rewrite significant parts of Zlog to maintain acceptable performance --
each such evolution costs developer time and risks introducing bugs.}

\newcommentone{We are actively exploring the use of high-level
\emph{declarative} languages based on Datalog~\cite{alvaro:cidr11} to program
storage APIs.  In this way, a programmer can specify the functional behavior of
a storage interface once in a relational (or algebraic) language, allowing an
optimizer to model search the space of functionally equivalent physical
implementations, and re-optimizing when characteristics change.  Much like
query planning and optimization in database systems~\cite{hellerstein:cidr15},
this approach will separate the concerns of correctness and performance, and
protect applications (which evolve slowly) against changes in highly dynamic
storage system.}
 \addressesissue{3}

@@

Another view of programmable storage systems is one of tailoring systems
resources to applications~\cite{arpaci:sosp01}. Related efforts include the
Exokernel~\cite{engler:sosp95}, SPIN~\cite{bershad:sosp95} and
Vino~\cite{seltzer:osdi96} projects; the latter two addressed the ability of injecting
code into the kernel to specialize resource management. Another approach is to
pass hints between the different layers of the IO stack to bridge the semantic
gap between applications and
storage~\cite{arpaci:sosp01,sivathanu:fast03,mesnier:sosp11}.

@@

\section{Conclusion}
\label{conclusion-and-future-work}

@@

\oldcommentone{Future work will focus on constructing more interfaces to support a wide
variety of storage system services that can be configured on-the-fly in existing
systems. This work is one point along that path to producing general-purpose
storage systems that can
target  special-purpose applications.  Ultimately we want to utilize
declarative methods for expressing new services.}
